Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-01-27 papers/LOPe/appendix-a.tex
Completion for coauthors review of sections 2 and 3.

Otherwise, Fortran syntax is usually straightforward though it may appear strange to the uninitiated
programmer.  A subroutine is a function with no return value and must be called in a statement
rather than as an expression.  Simple syntatic elements are: \texttt{!} the line comment symbol;
\texttt{\&} the line continuation symbol; and \texttt{::} is syntatic sugar used for the separation
of elements, e.g., separating type attributes from the variable list in a type declaration
statement.

>>>>

Otherwise, Fortran syntax is usually straightforward though it may appear strange to the uninitiated
programmer.  A subroutine is a function with no return value and must be called in a statement
rather than as an expression.  Simple syntatic elements are: \texttt{!} the line comment symbol;
\texttt{\&} the line continuation symbol; and \texttt{::} is syntatic sugar used for the separation
of syntatic elements, e.g., separating type attributes from the variable list in a type declaration
statement.

### 2015-01-27 papers/LOPe/coarray-comparison.tex
Completion for coauthors review of sections 2 and 3.

LOPe provides a purely \emph{local} viewpoint; the programmer is only provided read and write access
to the local array element and write access to a small halo region surrounding the local element.
There is simply \emph{no} mechanism provided for the programmer to even know \emph{where} the local
element is in the context of the broader array.  On a distributed memory architecture, the halo
elements may not even be physically located on the same processor.  If executed on a cluster
containing hybrid processing elements (e.g. GPUs), the halo elements may be as far as three hops
away: one to get to the host processor and another two to get to memory on the hybrid processor
executing on another distributed memory node.  LOPe provides a complete separation between algorithm
development and memory management and synchronization (between memory copies of the same logical
array region covered by halos).  By explicitly describing the existance and size of an array's halo
region, the compiler is provided with enough information to manage most of the hard and detailed
work involved in memory transfer and synchronization.  Additionally, the semantics of the LOPe
execution model remove the possibility of race conditions developing during execution of a
concurrent procedure.

We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
implementation of the same algorithm from the original Numrich and Reid paper describing coarrays in
Fortran.  We should point out that this comparison is somewhat unfair, because Numrich and Reid were
introducing coarray notation for transferring memory on distributed memory architectures, not
demonstrating how ideally to use coarrays within a large application.
%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
However this example serves
to highlight some of the advantages of LOPe and CAFe as introduced above.  Note that in the coarray
example shown below, type declarations have been removed to save space:
{\small \begin{verbatim}
  subroutine Laplace (nrow,ncol,U)
    left = me-1      ! me refers to the current image
    if (me == 1) left = ncol
    right = me + 1
    if (me == ncol) right = 1
    call sync_all( [left,right] )   ! Wait for left and right
    new_u(1:nrow) = new_u(1:nrow) + u(1:nrow)[left] + u(1:nrow)[right]
    call sync_all( [left,right] )
    u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
  end subroutine
\end{verbatim}}

The advantages of LOPe and CAFe over this example are now described.  Please note that the following
is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and
LOPe only pertains to the halo pattern useful in stencil-based algorithms.

@@

\item
LOPe applies the transfer of halo memory across possibly multiple levels of memory with the LOPe
intrinsic \texttt{TRANSFER\_HALO} function.  Thus the LOPe algorithm can be run on a machine with
many interconnected nodes, each containing hybrid processor cores.  The coarray example can only be
run on multiple nodes (called images in Fortran) without accelerator cores.

\item
Algorithm implementation is separate from user-specified synchronization, e.g., \texttt{call
  sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the \texttt{CONCURRENT}
attribute and the \texttt{TRANSFER\_HALO}.

@@

\item
The LOPe implementation is easier to understand and frequently follows the mathematical
algorithm directly.  For example, the CAF example adds 4 neighbors plus the center value
to make the implementation with direct remote coarray access possible, while the LOPe
example is able to implement the same algorithm with fewer operations by adding 4
neighbors (not including the center array element) and then only subtracting 3 center
values.

@@

\item
A programmer is not able to cause race conditions by forgetting to create and use temporary arrays
properly.  In LOPe it is the comilers responsibility to store data in temporary memory.

>>>>

LOPe provides a purely \emph{local} viewpoint; the programmer is only provided read and write access
to the local array element and write access to a small halo region surrounding the local element.
There is simply \emph{no} mechanism provided for the programmer to even know \emph{where} the local
element is in the context of the broader array.  On a distributed memory architecture, the halo
elements may not even be physically located on the same processor.  If executed on a cluster
containing hybrid processing elements (e.g. GPUs), the halo elements may be as far as three hops
away: one to get to the host processor and another two to get to memory on the hybrid processor
executing on another distributed memory node.  LOPe provides a complete separation between algorithm
development and memory management (synchronization between memory copies of the same logical
array region covered by halos).  By explicitly describing the existance and size of an array's halo
region, the compiler is provided with enough information to manage most of the hard and detailed
work involved in memory transfer and synchronization.  Additionally, the semantics of the LOPe
execution model remove the possibility of race conditions developing during execution of a
concurrent procedure.

We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
implementation of the same algorithm from the original Numrich and Reid paper first describing
coarrays in Fortran.  We should point out that this comparison is somewhat unfair, because Numrich
and Reid were introducing coarray notation for transferring memory on distributed memory
architectures, not demonstrating how ideally one should use coarrays within a large application.
%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
However this example serves
to highlight some of the advantages of LOPe and CAFe as introduced above.  Note that in the coarray
example shown below, type declarations have been removed to save space:
{\small \begin{verbatim}
  subroutine Laplace (nrow,ncol,U)
    left = me-1      ! me refers to the current image
    if (me == 1) left = ncol
    right = me + 1
    if (me == ncol) right = 1
    call sync_all( [left,right] )   ! Wait for left and right
    new_u(1:nrow) = new_u(1:nrow) + u(1:nrow)[left] + u(1:nrow)[right]
    call sync_all( [left,right] )
    u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
  end subroutine
\end{verbatim}}

The advantages of LOPe and CAFe over this example are now described.  Please note that the following
is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and
LOPe only pertains to the halo pattern most useful in the implementation of stencil-based algorithms.

@@

\item
LOPe applies the transfer of halo memory across possibly multiple levels of memory with the LOPe
intrinsic \texttt{TRANSFER\_HALO} function.  Thus the LOPe algorithm can be run on a machine with
many interconnected nodes, each containing hybrid processor cores.  The coarray example can only be
run on multiple nodes without accelerator cores.

\item
Algorithm implementation is separate from user-specified synchronization, e.g., \texttt{call
  sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the \texttt{CONCURRENT}
attribute and the \texttt{TRANSFER\_HALO} function call.

@@

\item
The LOPe implementation is easier to understand and frequently follows the mathematical algorithm
directly.  For example, the CAF implementation of Numrich and Reid adds 4 neighbors plus the center
value to make the implementation with direct remote coarray access possible, while the LOPe example
is able to implement the same algorithm with one statement and no intervening synchronization.

@@

\item
A programmer is not able to cause race conditions by forgetting to create and use temporary arrays
properly.  In LOPe it is the compilers responsibility to store data in temporary memory.

### 2015-01-27 papers/LOPe/coarray-extensions.tex
Completion for coauthors review of sections 2 and 3.

Consider the \texttt{Laplace} concurrent function discussed above.  In this section we demonstrate
how this function can be called in a context that allows full access to all of the interior elements
of the array, as well as array elements within the logically exterior, halo boundary region.  Topics
highlighted in this section are:
1. Distributed memory array allocation;
2. Explicit memory placement;
3. Remote memory transfer; and
4. Remote execution.
This description is within the context of a Domain Specific Language (DSL), again extending Fortran
2008.  As shorthand, these specific extensions are referred to as CAFe, for Coarray Fortran extensions.

@@

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1),          &
        dimension(:,:), codimensions(:,:) :: U
\end{verbatim}
with an explicit halo size and with two local dimensions (rank) and two distributed memory
codimensions (corank).  The corank of the array is chosen to be identical to the rank of
the array so that the logical process topology aligns is a way that allows a natural
halo exchange between logically neighboring processes.  For example, if my process
location is \texttt{[pcol,prow]}, then the right-hand halo for my local array \texttt{U}
can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size
and cosize of \texttt{U} are given by the allocation statement,
\begin{verbatim}
  allocate( U(0:M+1,0:N+1)[MP,*] )
\end{verbatim}

This allocation statement specifies (given the one element halo size provided earlier for
\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the
bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the enterior region
\texttt{U(1:M,1:N)}.  In this allocation statement, the total number of process rows \texttt{NP} can
be obtained at runtime, but may not be explicitly provided (according to CAF rules) because the
actual number of participating processes (in Fortran called images) is variable, depending on how
many processes are requested at startup.  In this discussion, it is \emph{assumed} that there are no
holes in the logical processor topology, thus $MP*NP = P$, where $P$ is the total number of
participating processes (images).

@@

One obtains a subimage by executing a function call,
\begin{verbatim}
   device = get_subimage(1)
\end{verbatim}
where the integer argument representing one of attached hardware devices.  If the function
fails (e.g., the requested device is unavailable) it returns the image number,
\texttt{this\_image()}, of the process that is executing the current program.  Returning
the current image allows program execution to proceed correctly even if there are no
attached devices.

@@

Fortran uses square bracket notation \texttt{[image]} to specify on what process the
memory reference is physically located.  Square brackets are a visual clue to the
programmer that the memory reference may be remote and therefore potentially suffer a
performance penalty.  CAFe extends this by employing double-bracket notation to indicate
possibly \emph{remote subimage execution}.

Execution of the \texttt{Laplace} task (defined previously) is done using the Fortran \texttt{do}
\texttt{concurrent} construct as shown,
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace( U(i,j)[device] )
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The local element of the array (as
defined above in reference to the definition of the concurrent procedure
\texttt{Laplace}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
execution of the loop body before further execution of the program proceeds; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
complete the halo transfer before program execution continues.

>>>>

Consider the \texttt{Laplace} concurrent function discussed above.  In this section we demonstrate
how this function can be called in the normal context of a program, one that allows full access to
all of the interior elements of the array, as well as array elements within the logically exterior,
halo-boundary region.  Topics highlighted in this section are: 1. Distributed memory array
allocation; 2. Explicit memory placement; 3. Remote memory transfer; and 4. Remote execution.  This
description is within the context of a Domain Specific Language (DSL), again extending Fortran 2008.
As shorthand, these specific extensions are referred to as CAFe, for Coarray Fortran extensions.

@@

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1),          &
        dimension(:,:), codimension(:,:) :: U
\end{verbatim}
with an explicit halo size and with two local dimensions (rank) and two distributed memory
codimensions (corank).  The corank of the array is chosen to be identical to the rank of
the array so that the logical process topology aligns in a way that allows a natural
halo exchange between logically neighboring processes.  For example, if my process
location is \texttt{[pcol,prow]}, then the right-hand halo for my local array \texttt{U}
can be obtained by the assignment \texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size
and cosize of \texttt{U} are given by the allocation statement,
\begin{verbatim}
  allocate( U(0:M+1,0:N+1)[MP,*] )
\end{verbatim}

This allocation statement specifies (given the one element halo size provided earlier for
\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the
bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the interior region
\texttt{U(1:M,1:N)}.  In this allocation statement, the total number of process rows \texttt{NP} can
be obtained at runtime, but \emph{may not} be explicitly provided (according to CAF rules) because the
actual number of participating processes (in Fortran called images) is variable, depending on how
many processes are requested at startup.  In this discussion, it is \emph{assumed} that there are no
holes in the logical processor topology, thus $MP*NP = P$, where $P$ is the total number of
participating processes (images).

@@

One obtains a subimage by executing a function call,
\begin{verbatim}
   device = get_subimage(1)
\end{verbatim}
where the integer argument representing one of attached hardware devices.  If the function
fails (e.g., the requested device is unavailable) it returns the image number
\texttt{this\_image()} of the process that is executing the current program.  Returning
the current image allows program execution to proceed correctly even if there are no
attached devices.

@@

Fortran uses square bracket notation, e.g. \texttt{[image]}, to specify on what process the
memory reference is physically located.  Square brackets are a visual clue to the
programmer that the memory reference may be remote and therefore potentially suffer a
performance penalty.  CAFe extends this by employing double-bracket notation to indicate
possibly \emph{remote subimage execution}.

Execution of the \texttt{Laplace} task is done using the Fortran \texttt{do}
\texttt{concurrent} construct:
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace( U(i,j)[device] )
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The local element of the array (as
defined above in reference to the definition of the concurrent procedure
\texttt{Laplace}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
execution of the loop body before further execution of the program proceeds; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
complete the halo transfer before program execution continues.

### 2015-01-27 papers/LOPe/local.tex
Completion for coauthors review of sections 2 and 3.

\author{Craig Rasmussen%
\thanks{The authors would like to thank Robert Robey and Wayne Weseloh
at Los Alamos National Laborabory for several stimulating
conversations with respect to programming models.}%
\and Matthew Sottile\and Dan Nagle\and Soren Rasmussen}

@@

\subsection{Acknowledgments}
This work was supported in part by the Department of Energy Office of Science,
Advanced Scientific Computing Research.

>>>>

\author{Craig Rasmussen
\and Matthew Sottile\and Dan Nagle\and Soren Rasmussen}

@@

\section*{Acknowledgments}
This work was supported in part by the Department of Energy Office of Science, Advanced Scientific
Computing Research.  The authors would also like to thank Robert Robey and Wayne Weseloh at Los
Alamos National Laborabory for several stimulating conversations with respect to programming models.

### 2015-01-27 papers/LOPe/programming-model.tex
Completion for coauthors review of sections 2 and 3.

The LOPe programming model (for Locally-Oriented Programming extensions) restricts the programmer to
a local view of the index space of an array.  Within a LOPe function, only a single array element (called
the local element) is mutable.  In addition, a small halo region surrounding the local element is
visible to the programmer, but this region is immutable.  Restricting the programmer to a local
index space serves to reduce complexity by removing all boundary-value and processor-topology
information from the implementation of the algorithm.

@@

LOPe was implemented as an embedded domain specific language (DSL).  Fortran was chosen as the base
language for LOPe, although in principle, the same techniques could be applied to language such as C
or C++ (in conjuction with some form of array extensions such as those proposed by Intel).  Fortran
was chosen because it already provides a rich array-based syntax.  Readers who are unfamiliar with
Fortran syntax may wish to consult Appendix A for a brief description of Fortran notation.

@@

\subsubsection{Halo regions.}

The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
an array boundary, so that an array may be referenced using indices that fall ``outside'' of the
domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler can
exploit this knowledge.  For example, a halo region can be declared with a statement of the form,
\begin{verbatim}
  real, allocatable, dimension(:), HALO(1:*:1) :: A
\end{verbatim}
This statement indicates that \texttt{A} is a rank one array, will be allocated later (when it will
be provided with a specific size), and has a halo region of one element surrounding the array on
either side.  Notation such as \texttt{M:*:N} is to be read as describing a halo region of size
\texttt{M} to the left, size \texttt{N} to the right, and with an arbitrary size in the middle
(denoted by the $*$ symbol).  When used to describe a formal parameter of a function, the size may
be ``assumed'' (denoted by a ``:'' symbol substituted in place of an explicit size) as is the
following,
\begin{verbatim}
  real, HALO(:,:) :: U
\end{verbatim}
which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
whose size is obtained by the compiler from the explicit halo size of the actual array argument
provided at the calling site of the function.  In LOPe, there is no need (in this instance) for a
repetitive \texttt{dimension(:,:)} specification as it is inferred from the \texttt{HALO}
specification.

\subsubsection{Concurrent functions.}

The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
of a \texttt{do} \texttt{concurrent} loop construct.  By using this construct, the programmer
asserts that specific iterations of the loop body may be executed by the compiler in \emph{any
  order;} even \emph{concurrently.}
%%LOPe allows the \texttt{concurrent} attribute to be attached to a function declaration so that the function can be called from within a concurrent loop without fear of race conditions.
For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  LOPe allows a function with the \texttt{pure} and \texttt{concurrent} attributes to be
called from within a \texttt{do} \texttt{concurrent} loop body.  One should imagine that this
function is called \emph{for each} \texttt{(i,j)} index of the interior of the array \texttt{U}.
(An example of this usage will be given later in the text.)  Note that this usage introduces a race
condition as new values of elements of \texttt{U} are created on the left-hand side of the
assignment statement that may use \emph{new or old} values of \texttt{U} on the right-hand side.
LOPe requires the compiler to guarantee that race conditions won't occur by using, e.g.,
double-buffering techniques as needed.

\subsubsection{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
array element and only the local element may be modified.  This zero-based indexing for the
local-array element is different than conventional Fortran, where by default, array indices start at
1.  For LOPe, this default convention is modified; indices for arrays declared with the
\texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in one
dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
\texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local, respectively) and
\texttt{U(0,-1)} and \texttt{U(0,+1)} (bottom and top of local, respectively).  The geometric
positioning of the array elements can be seen by examining the arrangement of the expressions on the
right-hand side of the \texttt{Laplace} example.

>>>>

The LOPe programming model (for Locally-Oriented Programming extensions) restricts the programmer to
a local view of the index space of an array.  Within a LOPe function, only a single array element (called
the local element) is mutable.  In addition, a small halo region surrounding the local element is
visible to the programmer, but this region is immutable.  Restricting the programmer to a local
index space serves to reduce complexity by separating all boundary-value and processor-topology
concerns from the implementation of the algorithm.

@@

LOPe was implemented as an embedded domain specific language (DSL).  Fortran was chosen as the base
language for LOPe, although in principle, the same techniques could be applied to language such as C
or C++ (in conjuction with some form of array extensions such as those proposed by Intel).  Fortran
was chosen because it already provides a rich array-based syntax.  Readers who are unfamiliar with
Fortran syntax may wish to consult Appendix A, for a brief description of Fortran notation.

@@

\textbf{Halo regions.}
The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
an array boundary, so that an array may be referenced using indices that fall ``outside'' of the
logical domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler
can exploit this knowledge.  For example, a halo region can be declared with a statement of the form,
\begin{verbatim}
  real, allocatable, dimension(:), HALO(1:*:1) :: A
\end{verbatim}
This statement indicates that \texttt{A} is a rank one array, will be allocated later (when it will
be provided with a specific size), and has a halo region of one element surrounding the array on
either side.  Notation such as \texttt{M:*:N} is to be read as describing a halo region of
\texttt{M} elements to the left, \texttt{N} elements to the right, and with an arbitrary number of
elements in the middle (denoted by the $*$ symbol).  When used to describe a formal parameter of a
function, the halo size may be ``assumed'' (denoted by a ``:'' symbol substituted in place of an
explicit size) as is the following,
\begin{verbatim}
  real, HALO(:,:) :: U
\end{verbatim}
which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
whose size is obtained by the compiler from the explicit halo size of the actual array argument
provided at the calling site of the function.  In LOPe, there is no need (in this instance) for a
repetitive \texttt{dimension(:,:)} specification, as it is inferred from the \texttt{HALO}
specification.

\textbf{Concurrent functions.}
The second keyword employed by LOPe is \texttt{concurrent}.  This keyword already exists in the form
of a \texttt{do} \texttt{concurrent} loop construct.  By using this construct, the programmer
asserts that specific iterations of the loop body may be executed by the compiler in \emph{any
  order;} even \emph{concurrently.}
%%LOPe allows the \texttt{concurrent} attribute to be attached to a function declaration so that the function can be called from within a concurrent loop without fear of race conditions.
For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  LOPe allows a function with the \texttt{pure} and \texttt{concurrent} attributes to be
called from within a \texttt{do} \texttt{concurrent} loop body.  One should imagine that this
function is called \emph{for each} \texttt{(i,j)} index of the interior of the array \texttt{U}.
(An example of this usage will be given later in the text.)  Note that this usage introduces a race
condition as new values of elements of \texttt{U} are created on the left-hand side of the
assignment statement that may use \emph{new or old} values of \texttt{U} on the right-hand side.
LOPe requires the compiler to guarantee that race conditions won't occur by using, e.g.,
double-buffering techniques as needed.

\textbf{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
array element and only the local element may be modified.  This zero-based indexing for the
local-array element is different than conventional Fortran, where by default, array indices start at
1.  For LOPe, this default convention is modified; indices for arrays declared with the
\texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in one
dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
\texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local, respectively) and
\texttt{U(0,-1)} and \texttt{U(0,+1)} (below and above of local, respectively).  The geometric
positioning of the array elements can be seen by examining the arrangement of the expressions on the
right-hand side of the \texttt{Laplace} example.
